Paper 3 - Causal Roadmap

### 2023-11-16 docs/causal_roadmap.qmd
Updated causal roadmap even if not using it for this paper.

---
title: "Paper 3 - Causal Roadmap"
keywords:
  - Causal Inference
  - Epidemiology
  - Triangulation
  - Modified Treatment Policy
  - Mendelian Randomization
  - Metabolomics
  - Endocrine Disrupting Chemicals
  - Neurodevelopment
  - Neurobehavior
metadata-files:
  - ../../../all_phd/authors.yml

format:
  pdf:
    pdf-engine: pdflatex
    number-sections: true
    toc: false
    margin-left: 15mm
    margin-right: 15mm
    margin-bottom: 20mm
    margin-top: 15mm
    include-in-header:
      text: |
        \usepackage{rotating}
        \usepackage{float}
        \usepackage{typearea}

bibliography: ../../../PhD.bib
link-citations: true
linkcolor: blue

execute:
  echo: false
  warning: false
  error: false
  message: false
editor_options: 
  chunk_output_type: console

filters:
  - ../../../all_phd/_extensions/acronyms/parse-acronyms.lua
  - ../../../all_phd/_extensions/authors-block/authors-block.lua
acronyms:
  fromfile: ../../../all_phd/acronyms.yml
  insert_links: false
  insert_loa: false
  sorting: alphabetical
---

@@

<!-- ####################################################################### -->
## Descriptive analyses
<!-- ####################################################################### -->

{{< include ../../../all_phd/descriptive_stats.qmd >}}

@@

```{r}
#| label: fig-causal-graph
#| fig.cap: Causal graph corresponding to the structural equations.
#| fig.height: 3.0
#| fig.width: 3.0
causal_graph <- ggdag::dagify(A ~ W + U_A, Y ~ W + A + U_Y, W ~ U_W, 
                              exposure = "A", outcome = "Y") %>%
  ggdag::tidy_dagitty()
ggdag::ggdag(causal_graph) + ggdag::theme_dag()
```

@@

```{r}
#| fig.cap: Simplified DAG for research question 1.
sdag1 <- ggdag::dagify(X_ti ~ W_t, 
                       Y ~ W_t + X_ti, 
                       exposure = "X_ti", outcome = "Y") %>%
  ggdag::tidy_dagitty() %>%
  ggdag::ggdag(text_size = 2, node_size = 10) + ggdag::theme_dag() +
  ggplot2::ggtitle("RQ1")
```

@@

```{r}
#| fig-cap: Simplified DAG for MR analyses of `outcome ~ omic` associations.
#| label: fig-dag-mr
#| fig-width: 5
#| fig-height: 4
coords <- tibble::tribble(
  ~name, ~x, ~y, 
  "neuro", 1, 0, 
  "metab", 0.5, 0, 
  "SNP", 0, 0, 
  "C", 0.75, 0.1
)
ggdag::dagify(
  neuro ~ metab + C, 
  metab ~ SNP + C, 
  exposure = "metab", 
  outcome = "neuro", 
  coords = coords
) %>%
  ggdag::ggdag(., layout = "circle") +
  ggdag::geom_dag_point(col = "gray") +
  ggdag::geom_dag_edges() +
  ggdag::geom_dag_text(col = "black") +
  ggplot2::theme(axis.title = ggplot2::element_blank(), 
                 axis.ticks = ggplot2::element_blank(), 
                 axis.text = ggplot2::element_blank())
```

@@

## Reliable Causal Inference

{{< include ../../../all_phd/app-rel-ci.qmd >}}

## Reliable Statistical Inference

{{< include ../../../all_phd/app-rel-si.qmd >}}

## Replicability

{{< include ../../../all_phd/app-replicability.qmd >}}

@@

<!-- ####################################################################### -->
## Data Checking
<!-- ####################################################################### -->

{{< include ../../../all_phd/app-data-checking.qmd >}}

@@

| **Name** | **Version** | **Description** |
|:---------|:------------|:----------------|
| Random seed | NA | Will be set to $X$ |
| `R` | `r I(paste0(R.version$major, ".", R.version$minor))` | Statistical programming environment |
| `tlverse` | `r packageVersion("tlverse")` | Targeted Learning software ecosystem in R |
| `lmtp` | `r packageVersion("lmtp")` | Non-parametric Causal Effects of Feasible Interventions Based on Modified Treatment Policies |
| `sl3` | `r packageVersion("sl3")` | Super Learner ensemble machine learning |
| `labelled` | `r packageVersion("labelled")` | Manipulate metadata |
| `dlookr` | `r packageVersion("dlookr")` | Diagnose, explore, and transform data |
| `Hmisc` | `r packageVersion("Hmisc")` | Harrell Miscellaneous |
| `dagitty` | `r packageVersion("dagitty")` | Graphical Analysis of Structural Causal Models |
| `qreport` | `r packageVersion("qreport")` | Facilitating the composition of Quarto reports |
| ... | ... | ... |

@@

<!-- ####################################################################### -->
## Gene-Environment Interactions
<!-- ####################################################################### -->

Genetic polymorphisms relevant for the **chemicals**:

* \acr{opps}:
  * The \acr{snps} of the *PON1* (paraoxonase 1) gene, known to influence the ability to metabolize and detoxify \acr{opps}: `rs854560` and `rs662`. [Reference](https://doi.org/10.1289/ehp.1408976).
  * The \acr{snps} of the genes coding for \acr{opps}-metabolizing enzymes: `rs1803274`, `rs1799807`, `rs662`, `rs854560`, `rs705379`, `rs4244285`, `rs3892097`, and `rs2177180`. [Reference](https://doi.org/10.1016/j.chemosphere.2015.07.079).
* Phenols:
  * The \acr{snp} of the *UGT2B15* gene: `rs1902023`. [Reference](https://doi.org/10.1016/j.envint.2020.106172).
* Phthalates:
  *  The \acr{snps} of the cytochrome P450 (*CYPs*) and UDP-glucuronosyltransferase (*UGTs*) genes: `rs1057910`, `rs1799853`, `rs4244285`, `rs12248560`, `rs3892097`, `rs1902023`, and `rs11692021`. [Reference](https://doi.org/10.1016/j.envint.2021.107046).
  * The \acr{snps} of the *PPAR*$\alpha$ gene: `rs75525202`, `rs4823902`, `rs4253690`, `rs12330015`, `rs5766698`, `rs115250492`, and `rs1800246`. [Reference](https://doi.org/10.3389/fgene.2022.855544). Not sure whether these \acr{snps} should be considered in the list below.

Genetic polymorphisms relevant for the **outcomes**:

>>>>

---
title: "Paper 3 - Causal Roadmap"
keywords:
  - Causal Inference
  - Epidemiology
  - Triangulation
  - Modified Treatment Policy
  - Mendelian Randomization
  - Metabolomics
  - Endocrine Disrupting Chemicals
  - Neurodevelopment
  - Neurobehavior
format:
  pdf:
    pdf-engine: pdflatex
    number-sections: true
    toc: false
    margin-left: 15mm
    margin-right: 15mm
    margin-bottom: 20mm
    margin-top: 15mm
    include-in-header:
      text: |
        \usepackage{rotating}
        \usepackage{float}
        \usepackage{typearea}
bibliography: ../../../PhD.bib
link-citations: true
linkcolor: blue
execute:
  echo: false
  warning: false
  error: false
  message: false
editor_options: 
  chunk_output_type: console
filters:
  - ../../../all_phd/_extensions/acronyms/parse-acronyms.lua
acronyms:
  fromfile: ../../../all_phd/acronyms.yml
  insert_links: false
  insert_loa: false
  sorting: alphabetical
---

@@

@@

```{r}
#| label: fig-causal-graph
#| fig.cap: Causal graph corresponding to the structural equations.
#| fig.height: 3.0
#| fig.width: 3.0
causal_graph <- ggdag::dagify(A ~ W + U_A, Y ~ W + A + U_Y, W ~ U_W, 
                              exposure = "A", outcome = "Y") |>
  ggdag::tidy_dagitty()
ggdag::ggdag(causal_graph) + ggdag::theme_dag()
```

@@

```{r}
#| fig.cap: Simplified DAG for research question 1.
sdag1 <- ggdag::dagify(X_ti ~ W_t, 
                       Y ~ W_t + X_ti, 
                       exposure = "X_ti", outcome = "Y") |>
  ggdag::tidy_dagitty() |>
  ggdag::ggdag(text_size = 2, node_size = 10) + ggdag::theme_dag() +
  ggplot2::ggtitle("RQ1")
```

@@

```{r}
#| fig-cap: Simplified DAG for MR analyses of `outcome ~ omic` associations.
#| label: fig-dag-mr
#| fig-width: 5
#| fig-height: 4
coords <- tibble::tribble(
  ~name, ~x, ~y, 
  "neuro", 1, 0, 
  "metab", 0.5, 0, 
  "SNP", 0, 0, 
  "C", 0.75, 0.1
)
ggdag::dagify(
  neuro ~ metab + C, 
  metab ~ SNP + C, 
  exposure = "metab", 
  outcome = "neuro", 
  coords = coords
) |>
  ggdag::ggdag(layout = "circle") +
  ggdag::geom_dag_point(col = "gray") +
  ggdag::geom_dag_edges() +
  ggdag::geom_dag_text(col = "black") +
  ggplot2::theme(axis.title = ggplot2::element_blank(), 
                 axis.ticks = ggplot2::element_blank(), 
                 axis.text = ggplot2::element_blank())
```

@@

@@

@@

| **Name** | **Version** | **Description** |
|:---------|:------------|:----------------|
| Random seed | NA | Will be set to $X$ |
| `R` | `r I(paste0(R.version$major, ".", R.version$minor))` | Statistical programming environment |
| `lmtp` | `r packageVersion("lmtp")` | Non-parametric Causal Effects of Feasible Interventions Based on Modified Treatment Policies |
| `labelled` | `r packageVersion("labelled")` | Manipulate metadata |
| `dlookr` | `r packageVersion("dlookr")` | Diagnose, explore, and transform data |
| `dagitty` | `r packageVersion("dagitty")` | Graphical Analysis of Structural Causal Models |
| ... | ... | ... |

@@
